---
title: Analysing your own Strava data in R using rStrava
author: Daniel Padfield
date: '2018-03-03'
slug: analysing-your-own-strava-data-in-r-using-rstrava
draft: true
categories:
  - R
  - strava
tags: ["strava"]
---

Over the last 4 years I have consistently cycled around or over 2000 km a year around Cornwall. Almost all of these rides have been uploaded to Strava which has a suite of nice ways to visualise your performance and each ride.

However, the ride data on Strava is mine, I have literally created it, and as such I looked for a way of accessing all of that beautiful data. One way would be to download all your .gpx files and read them in. That is going to be a lot of data. The alternative I settled on is rStrava, an R package which wraps the Strava API in many easy-to-use functions allowing access to all your glorious data.

## Installing rStrava and creating a authorisation token

```{r setup, eval = FALSE}
# load packages ####
library(rStrava) # devtools::install_github('fawda123/rStrava')
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(ggridges)
library(GGally)

# initial setup ####
# Strava key
app_name <- 'xxxx'
app_client_id <- 'xxxxx'
app_secret <- '"xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx"'

# create strava token
my_token <- httr::config(token = strava_oauth(app_name, app_client_id, app_secret))

# Google elevation API key
GoogleAPI <- 'xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx'

```

```{r setup2, echo=FALSE, eval = FALSE}
# load packages ####
library(rStrava) # devtools::install_github('fawda123/rStrava')
library(dplyr)
library(tidyr)
library(purrr)

# initial setup ####
# Strava key
app_name <- 'stRava'
app_client_id <- '12962'
app_secret <- 'e58f0335795fb08674d2a412c9d35efb0d740761'

# Google elevation API key
GoogleAPI <- 'AIzaSyDgUnXCaOSQ90xZ9wUYGP0E9BUtIkm2gkU'

# create token
my_token <- httr::config(token = strava_oauth(app_name, app_client_id, app_secret))
```

## Downloading your data

```{r get_compile_data, eval = FALSE}
# download strava data
my_acts <- get_activity_list(my_token, id = 2140248)

# compile activities into a tidy dataframe
my_acts <- compile_activities(my_acts)

# have a look at the dataframe
dplyr::glimpse(my_acts)
```

Do some transformations.

```{r transform_and_pick, eval = FALSE}
desired_cols <- desired_columns <- c('achievement_count', 'athlete.id', 'average_cadence', 'average_heartrate', 'average_speed', 'average_watts', 'distance', 'elapsed_time', 'elev_high', 'elev_low', 'kilojoules', 'suffer_score', 'max_heartrate', 'max_speed', 'moving_time', 'start_date', 'start_date_local', 'name', 'total_elevation_gain', 'type', 'map.summary_polyline', 'location_city', 'upload_id')

# delete data with a distance < 3 km
my_acts <- filter(my_acts, distance > 3 & type != 'Workout')

# transformations of data
my_acts <- mutate_at(my_acts, c('average_cadence', 'average_heartrate', 'average_speed', 'average_watts', 'distance', 'elapsed_time', 'elev_high', 'elev_low', 'kilojoules', 'suffer_score', 'max_heartrate', 'max_speed', 'moving_time', 'total_elevation_gain'), as.numeric) %>%
  mutate(.,
         activity_no = seq(1,nrow(.), 1),
         elapsed_time = elapsed_time/60/60,
         moving_time = moving_time/60/60,
         elevation_per_km = total_elevation_gain/distance,
         date = gsub("T.*$", '', start_date) %>%
           as.POSIXct(., format = '%Y-%m-%d'),
         EUdate = format(date, '%d/%m/%Y'),
         month = format(date, "%m"),
         day = format(date, "%d"),
         year = format(date, "%Y")) %>%
  mutate_at(., c('month', 'day', 'year'), as.numeric) %>%
  arrange(., date) %>%
  group_by(year, type) %>%
  mutate(cum_distance = cumsum(distance)) %>%
  data.frame()


```

## The opportunities are endless

__1. Plot cumulative distance through time__

```{r plot_1, eval = FALSE}
ggplot(filter(my_acts, type == 'Ride'), aes(distance, as.character(month))) +
  geom_density_ridges() +
  facet_wrap(~year)
```

### Plot distance through time

### Plot speed against predictors

```{r pairs_plot, eval = FALSE}
ggpairs(filter(my_acts, type == "Ride" & distance > 20), 
        columns = c('average_speed', 'distance', 'average_heartrate', 'average_cadence', 'total_elevation_gain', 'suffer_score', 'elevation_per_km'),
        mapping = aes(col = as.character(year)))
```